---
title: "Northern Three Spatial Analyses (Estuarine: Water Quality)"
author: "Adam Shand"
date: "`r format(Sys.time(), '%d, %B, %Y')`"
format: html
params:
  project_crs: "EPSG:7844"
---

A simple script to create a map for the sampling locations/extent for estuarine water quality. 

# Global Controls

As always we begin by setting the global variables for the script, this includes loading packages, the output save location, the script crs, and a date-time tracker.

```{r}
#| label: Load packages
#| output: false
#| echo: false

#mapping packages
library(RColorBrewer) #colour palettes, and create custom palettes
library(sf) #read and manipulate simple features
library(terra)
library(tmap) #map data
library(readxl)

#general setup and data manipulation
library(glue) #easier to create strings, particularly those contains variables
library(lubridate) #time manipulation
library(tidyverse) #contains multiple useful packages (https://www.tidyverse.org/packages/)
library(dataaimsr) #load the aims datasets
```

```{r}
#| label: global vars and initial setup
#| output: false

#set project variables: crs factor and current_fyear
proj_crs <- params$project_crs

#format the date for the path below and to be used as a variable in naming later
date <- format(Sys.time(), "%Y-%m-%d")

#create a file path to help with saving things
save_path <- glue("output/{date}_n3-estuarine_water-quality-maps/")

#bring that path to life
dir.create(save_path)

#turn off s2 geometry
sf_use_s2(F)

```

# Load data

Next we load in the data. This data can be found in the **data/estuarine_water-quality** folder in the github repo. If this is the first time running the analysis please refer to the .README for this repo for a guide on accessing the data.

```{r}
#| label: load in wetland data
#| output: false

#please note if the data is not in the basin builder folder, try running the basin builder script
dry_tropics <- st_read("data/dry-tropics_basin-builder/Dry-Tropics-Basins-Detailed.shp") |> st_transform(proj_crs)

#read in qld outlines data from the gisaimsr package, filter for land and islands, update crs
qld <- get(data("gbr_feat", package = "gisaimsr")) |> filter(FEAT_NAME %in% c("Mainland", "Island")) |> 
  st_transform(proj_crs)

#read data and transform into shapefile
wq <- st_as_sf(read_excel("data/estuarine_water-quality/DT_estuarine_sites_gv_lor_sf_area_raw.xlsx",
                           sheet = "sites"), coords = c("long", "lat")) |> st_set_crs(proj_crs)

```

# Maipulate Data

Before we can create the map we just need to manipulate the data bit to remove unnecessary visual clutter.

```{r}
#| label: clean up data

#group the data for our region and drop everything but the estuarine sections
dry_tropics_est <- dry_tropics |> filter(env == "Estuarine") |> 
  rename(Basin = zone) |> 
  mutate(Basin = case_when(Basin == "black" ~ "Black",
                          Basin == "ross" ~ "Ross"))

#get only land areas for map background
dr_tropics_background <- dry_tropics |> filter(env != "marine") |> 
  group_by(zone) |> summarise(geometry = st_union(geometry))


#filter for only current
wq <- wq |> filter(current_site == "Yes")

#filter for only Ross data
wq_r <- wq |> filter(basin == "Ross")

#filter for only Black data
wq_b <- wq |> filter(basin == "Black")


```

# Create Map

The core objective of this map is to visually present the water sampling locations in our technical report style.

```{r}
#| label: create map
#| output: false

#create a townsville point coord
test1 <- data.frame(place = "Townsville", x = "-19.2610", y = "146.8129")
test2 <- st_as_sf(test1, coords = c("y", "x"), crs = proj_crs)

#extract ross only
dt_ross <- dry_tropics_est |> filter(Basin == "Ross")

#extract blac only
dt_black <- dry_tropics_est |> filter(Basin == "Black")


#create a map of the area
map <- tm_shape(qld) +
  tm_polygons(col = "grey80", border.col = "black") +
  tm_shape(dr_tropics_background) +
  tm_polygons(border.col = "black") +
  tm_shape(dt_ross) +
  tm_borders() +
  tm_shape(test2) +
  tm_symbols(size = 0.5, col = "white", border.col = "black", border.lwd = 2, shape = 23) +
  tm_text("place", shadow = T, xmod = -2.5, ymod = 0.1) +
  tm_shape(wq_r) +
  tm_symbols(size = 0.8, col = "watercourse", border.col = "black", border.lwd = 2, shape = 21) +
  tm_text("site", shadow = T, auto.placement = 0.08, size = 1) +
  tm_shape(st_buffer(wq_r, 0.05), is.master = T) +
  tm_symbols(size = 0) +
  tm_add_legend(type = "line", col = "grey60", labels = "Boundaries")  +
  tm_layout(legend.frame = T, legend.bg.color = "White", asp = 1.1, 
            legend.text.size = 0.7, legend.position = c("left", "bottom")) +
  tm_scale_bar(width = 0.15, text.size = 0.7, position = c(0.18, 0)) +
  tm_compass(position = c("right", "top"))

#save the map as a png
tmap_save(map, filename = glue("{save_path}estuarine_water-quality-ross.png"))

#create a map of the area
map <- tm_shape(qld) +
  tm_polygons(col = "grey80", border.col = "black") +
  tm_shape(dr_tropics_background) +
  tm_polygons(border.col = "black") +
  tm_shape(dt_black) +
  tm_borders() +
  tm_shape(test2) +
  tm_symbols(size = 0.5, col = "white", border.col = "black", border.lwd = 2, shape = 23) +
  tm_text("place", shadow = T, xmod = -2.5, ymod = 0.1) +
  tm_shape(wq_b) +
  tm_symbols(size = 0.8, col = "watercourse", border.col = "black", border.lwd = 2, shape = 21) +
  tm_text("site", shadow = T, auto.placement = 0.05, size = 1) +
  tm_shape(st_buffer(wq_b, 0.05), is.master = T) +
  tm_symbols(size = 0) +
  tm_add_legend(type = "line", col = "grey60", labels = "Boundaries")  +
  tm_layout(legend.frame = T, legend.bg.color = "White", asp = 1.1, 
            legend.text.size = 0.7, legend.position = c("left", "bottom")) +
  tm_scale_bar(width = 0.15, text.size = 0.7, position = c(0.21, 0)) +
  tm_compass(position = c("right", "top"))

#save the map as a png
tmap_save(map, filename = glue("{save_path}estuarine_water-quality-black.png"))



```


# Create Estuarine overview map

```{r}
#| label: inshore marine overview map

#create a map of the area
map <- tm_shape(qld) +
  tm_polygons(col = "grey80", border.col = "black") +
  tm_shape(dr_tropics_background) +
  tm_polygons(border.col = "black") +
  tm_shape(dry_tropics_est, is.master = T) +
  tm_polygons(col = "Basin", border.col = "black") +
  tm_shape(test2) +
  tm_symbols(size = 0.5, col = "white", border.col = "black", border.lwd = 2, shape = 23) +
  tm_text("place", shadow = T, xmod = 2.5, ymod = 0.1) +
  tm_layout(legend.frame = T, legend.bg.color = "White", asp = 1.1, 
            legend.text.size = 0.7, legend.position = c("left", "bottom")) +
  tm_scale_bar(width = 0.15, text.size = 0.7, position = c(0.12, 0)) +
  tm_compass(position = c("right", "top"))

#save the map as a png
tmap_save(map, filename = glue("{save_path}estuarine_zones.png"))

```
