---
title: "Inshore Marine Water Quality Exploratory Data Analysis"
subtitle: "A Healthy Waters Partnership Analysis"
description: "Script 4 in a series of script designed to analyse, score, and present inshore marine water quality in the Dry Tropics region. The output of this is used in the Dry Tropics Technical Report."
author: "Adam Shand"
format: html
params:
  project_crs: "EPSG:7844"
  target_fyear: 2023 
---

:::{.callout-note}
The Water Quality suite of scripts currently do not adhere to the CamelCase naming rules. This is due to filtering and code that relies on snake_case naming to work. This will take a significant amount of time to overhaul.
:::

# Introduction

The purpose of this script is to create maps identifying the sampling site locations for the inshore marine water quality index.

These maps are used in the body of the methods document, and in the appendix of the technical report.

# Script Set Up

This script requires multiple core spatial packages, each of which is loaded below.

```{r}
#| label: load packages

#use pacman function to load and install (if required) all other packages
pacman::p_load(tidyverse, glue, here, sf, tmap, janitor, readxl, grid, RColorBrewer)

```

We also need to set up key variables for the script such as where we are going to save the outputs, and what coordinate reference system (crs) we are going to be using.

```{r}
#| label: global vars and initial setup
#| output: false

#set project variables
proj_crs <- params$project_crs
current_fyear <- params$target_fyear

#get a date variable (this is important for naming as it is anticipated multiple runs of the script will be required).
date <- format(Sys.time(), "%Y-%m-%d")

#create a file path for the year of data that is being looked at
year_folder <- glue("{here()}/outputs/dt_water-quality_inshore_s4-maps/{current_fyear-1}-{current_fyear}_maps/")

#create that folder
dir.create(year_folder, recursive = T)

#create a file path to help with saving things, make sure to include date
save_path <- glue("{year_folder}/maps_created_on_{date}/")

#create folder
dir.create(save_path, recursive = T)

#turn off spherical geometry
sf_use_s2(F)

```

# Load Data

Now the script is set up we need to load in all of the required datasets. Spatial data for the northern three regions should be readily available in the repo, the dataset is created by the n3_region-builder.qmd script in the repo that should be the first script run for new users. If the dataset is not available refer back to the README document in the GitHub repo. The other parts of spatial data here should go off without a hitch if the region-builder script has been run.

```{r}
#| label: load in all data
#| output: false

#read in the northern three spatial files and reduce down to only the components we need
dt_marine <- st_read(here("data/n3_prep_region-builder/n3_region.gpkg")) |> 
  filter(region == "Dry Tropics",
         !str_detect(basin_or_zone, "Burd|Haug|Off"),
         environment == "Marine",
         is.na(stream_order)) |> 
  mutate(geographic_area = case_when(geographic_area == "Open Coastal" ~ "H.Open Coastal",
                                     geographic_area == "Enclosed Coastal" ~ "H.Enclosed Coastal",
                                     T ~ geographic_area))

#get land components to help with background
dt_land <- st_read(here("data/n3_prep_region-builder/n3_region.gpkg")) |> 
  filter(region == "Dry Tropics",
         !str_detect(basin_or_zone, "Burd|Haug"),
         environment != "Marine",
         is.na(stream_order)) |> 
  group_by(basin_or_zone) |> 
  summarise(geom = st_union(geom)) |> 
  ungroup() |> st_cast() |> st_make_valid()

#load a Queensland outline from the aims package
qld <- get(data("gbr_feat", package = "gisaimsr")) |> filter(FEAT_NAME %in% c("Mainland", "Island")) |> 
  st_transform(proj_crs)

#read in site coords and convert to a simple feature object
sites <- read_excel(path = here("data/dt_water-quality_inshore/raw/dt_wq_inshore_metadata.xlsx"), sheet = "Current_Sites") |> 
  select(Zone, Sub_Zone, Geographic_Area, Code, Lat, Long) |> 
  st_as_sf(coords = c("Long", "Lat"), crs = proj_crs)

#calculate extent of query area
dt_bbox <- st_bbox(dt_land)

#query the osm database for waterways and natural waters (no idea why they are separate categories) and points of interest
wat <- opq(dt_bbox, timeout = 100) |> add_osm_feature(key = "waterway", value = c("river", "stream")) |> osmdata_sf()
nat <- opq(dt_bbox, timeout = 100) |> add_osm_feature(key = "natural", value = "water") |> osmdata_sf()
poi <- opq(dt_bbox, timeout = 100) |> add_osm_feature(key = "place", value = c("town", "city")) |> osmdata_sf()

#transform and filter data to only get specific data types
wat <- st_transform(wat$osm_lines, crs = proj_crs)
nat <- st_transform(nat$osm_polygons, crs = proj_crs)
poi <- st_transform(poi$osm_points, crs = proj_crs)

```

# Create Map

Now we can put everything together. We will loop over the Ross and Black to make individual maps.

```{r}
#| label: create region maps

#create target vector to loop lover
target_vect <- unique(dt_marine$basin_or_zone)

#create a similar vect for land basins
land_vect <- rev(unique(dt_land$basin_or_zone))

for (i in 1:length(target_vect)){
  
  #filter sites and rename column
  filtered_sites <- sites |> filter(Zone == target_vect[i]) |> rename("Geographic Area: Reported" = Geographic_Area)
  
  #define focus area for waterways and rename column
  focus_area <- dt_land |> filter(basin_or_zone == land_vect[i]) #|> rename("Sub Basin: All" = sub_basin_or_sub_zone)
  
  #define focus area for marine wq sites and rename column
  focus_marine <- dt_marine |> filter(basin_or_zone == target_vect[i]) |> rename("Geographic Area: All" = geographic_area)
  
  #create a box of the focus area
  focus_bbox <- st_as_sfc(st_bbox(focus_marine))
  
  #crop waterways to be within focus area
  focus_wat <- wat |> st_intersection(st_union(focus_area))
  focus_nat <- nat |> st_intersection(st_union(focus_area))
  
  #create a secondary focus area
  focus2_bbox <- st_buffer(st_as_sfc(st_bbox(filtered_sites)), 0.01)
  
  #create an inset map
  inset_map <- tm_shape(qld) +
    tm_polygons(col = "grey80", border.col = "black") +
    tm_shape(dt_marine, is.master = T) +
    tm_polygons(col = "grey90", border.col = "black") +
    tm_shape(dt_land) +
    tm_polygons(col = "grey90", border.col = "black") +
    tm_shape(focus_bbox) +
    tm_borders(lwd = 2, col = "red") +
    tm_shape(focus2_bbox) +
    tm_borders(lwd = 2, col = "blue")
  
  #create the main map
  map <- tm_shape(focus2_bbox, is.master = T) +
    tm_polygons(alpha = 0) +
    tm_shape(qld) +
    tm_polygons(col = "grey80", border.col = "black") +
    tm_shape(dt_marine) +
    tm_polygons(col = "white", border.col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(col = "grey90", border.col = "grey90") +
    tm_shape(dt_land) +
    tm_polygons(col = "grey90", border.col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(col = "Geographic Area: All", border.col = "black", alpha = 0.5, palette = "Pastel1") +
    tm_shape(focus_wat) +
    tm_lines(col = "dodgerblue") +
    tm_shape(focus_nat) +
    tm_polygons(col = "aliceblue", border.col = "dodgerblue") +
    tm_shape(poi) +
    tm_symbols(size = 0.3, col = "white", border.col = "black", border.lwd = 1, shape = 23) +
    tm_text("name", shadow = T, auto.placement = T, size = 0.8) +
    tm_shape(filtered_sites) +
    tm_symbols(size = 0.6, col = "Geographic Area: Reported", border.col = "black", border.lwd = 2, shape = 21, palette = "Set1") +
    tm_text("Code", shadow = T, auto.placement = T, size = 0.9) +
    tm_layout(legend.bg.color = "white", legend.frame = "black", asp = 1.1, legend.position = c("right", "bottom"))
  
  #figure out the aspect of the inset map and the view port
  asp2 <- (dt_bbox$ymax - dt_bbox$ymin)/(dt_bbox$xmax - dt_bbox$xmin)
  w <- 0.2
  h <- asp2 * w
  vp <- viewport(x = 0.98, y = 0.97, width = w, height = h, just = c("right", "top"))

  #save map
  tmap_save(map, glue("{save_path}/{target_vect[i]}_Inshore.png"), insets_tm = inset_map, insets_vp = vp)
  
}

```

# Create De-identified Map

Below we will create the same maps, but now with sites de-identified.

```{r}
#| label: create deidentified region maps

#create target vector to loop lover
target_vect <- unique(dt_marine$basin_or_zone)

#create a similar vect for land basins
land_vect <- rev(unique(dt_land$basin_or_zone))

for (i in 1:length(target_vect)){
  
  #filter sites and rename column
  filtered_sites <- sites |> filter(Zone == target_vect[i]) |> rename("Geographic Area: Reported" = Geographic_Area)
  
  #define focus area for waterways and rename column
  focus_area <- dt_land |> filter(basin_or_zone == land_vect[i]) #|> rename("Sub Basin: All" = sub_basin_or_sub_zone)
  
  #define focus area for marine wq sites and rename column
  focus_marine <- dt_marine |> filter(basin_or_zone == target_vect[i]) |> rename("Geographic Area: All" = geographic_area)
  
  #create a box of the focus area
  focus_bbox <- st_as_sfc(st_bbox(focus_marine))
  
  #crop waterways to be within focus area
  focus_wat <- wat |> st_intersection(st_union(focus_area))
  focus_nat <- nat |> st_intersection(st_union(focus_area))
  
  #create a secondary focus area
  focus2_bbox <- st_buffer(st_as_sfc(st_bbox(filtered_sites)), 0.01)
  
  #create an inset map
  inset_map <- tm_shape(qld) +
    tm_polygons(col = "grey80", border.col = "black") +
    tm_shape(dt_marine, is.master = T) +
    tm_polygons(col = "grey90", border.col = "black") +
    tm_shape(dt_land) +
    tm_polygons(col = "grey90", border.col = "black") +
    tm_shape(focus_bbox) +
    tm_borders(lwd = 2, col = "red") +
    tm_shape(focus2_bbox) +
    tm_borders(lwd = 2, col = "blue")
  
  #create the main map
  map <- tm_shape(focus2_bbox, is.master = T) +
    tm_polygons(alpha = 0) +
    tm_shape(qld) +
    tm_polygons(col = "grey80", border.col = "black") +
    tm_shape(dt_marine) +
    tm_polygons(col = "white", border.col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(col = "grey90", border.col = "grey90") +
    tm_shape(dt_land) +
    tm_polygons(col = "grey90", border.col = "black") +
    tm_shape(focus_marine) +
    tm_polygons(col = "Geographic Area: All", border.col = "black", alpha = 0.5, palette = "Pastel1") +
    tm_shape(focus_wat) +
    tm_lines(col = "dodgerblue") +
    tm_shape(focus_nat) +
    tm_polygons(col = "aliceblue", border.col = "dodgerblue") +
    tm_shape(poi) +
    tm_symbols(size = 0.3, col = "white", border.col = "black", border.lwd = 1, shape = 23) +
    tm_text("name", shadow = T, auto.placement = T, size = 0.8) +
    tm_shape(filtered_sites) +
    tm_symbols(size = 0.6, col = "Geographic Area: Reported", border.col = "black", border.lwd = 2, shape = 21, palette = "Set1") +
    #tm_text("Code", shadow = T, auto.placement = T, size = 0.9) +
    tm_layout(legend.bg.color = "white", legend.frame = "black", asp = 1.1, legend.position = c("right", "bottom"))
  
  #figure out the aspect of the inset map and the view port
  asp2 <- (dt_bbox$ymax - dt_bbox$ymin)/(dt_bbox$xmax - dt_bbox$xmin)
  w <- 0.2
  h <- asp2 * w
  vp <- viewport(x = 0.98, y = 0.97, width = w, height = h, just = c("right", "top"))

  #save map
  tmap_save(map, glue("{save_path}/{target_vect[i]}_de-identified_Inshore.png"), insets_tm = inset_map, insets_vp = vp)
  
}

```

